Learning by Teaching a Robot:\\The Case of Handwriting

### 2015-12-17 main.tex
Large proof-reading

\author{Séverin Lemaignan$^1$, Alexis Jacq$^{1,2}$, Fernando Garcia$^1$,
    Deanna Hood$^1$, \\Aude
    Billard$^1$, Ana Paiva$^2$, Pierre Dillenbourg$^1$ \\
$^1$CHILI Lab, École Polytechnique Fédérale de Lausanne, Suisse,\\
$^2$Instituto Superior Técnico, University of Lisbon, Portugal}

@@

Thomas is five and half, and has been diagnosed with visuo-constructive
deficits. He is under the care of an occupational therapist, and tries to
workaround his inability to draw letters in a consistent manner. Vincent is six
and struggles at school with his poor handwriting and even poorer
self-confidence.\footnote{Children's names have been changed.}

@@

Remediations for handwriting difficulties traditionally involve long
interventions (at least 10 weeks~\cite{Hoy2011}), essentially consisting in
handwriting training with occupational therapists, and primarily addressing the
\emph{motor} deficits.  Improvements in self-confidence and anxiety occur (at
best) as a side-effect of the child improving his/her handwriting skills and,
consequently, improving his/her performance at school.

We present in this article a new take on this educative challenge: a remediation
procedure that involve a ``bad writer'' robot that is \emph{taught} by the
child. By building on the \emph{learning by teaching} paradigm, not only the
child practises handwriting but, as (s)he takes on the role of the teacher,
(s)he also positively reinforce his/her self-esteem and motivation: his/her
social role shifts from the ``underperformer'' to ``the one who knows and
teaches''. And by relying on a robot, we can tailor the exercises and the
learning rate to each children' needs, as we show in this article.

\subsection{Learning by Teaching}

@@

We also investigate here a particular role for a robot in the
education of handwriting: not only is the robot actively performing the activity
by drawing letters, but it does so in a way that engages the child in a very
specific social role. The child is the teacher in this relationship and the
robot is the learner: the child must engage in a (meta-) cognitive relationship
with the robot to try to understand why the robot fails and how to help it best.
Here, the robot is more than just an activity facilitator or orchestrator -- its
physical presence and embodiment induce agency and anthropomorphising, and
cognitively engage the child into the learning activity (be it consciously or
not).

Besides the commitment of the child into the interaction build on a
psychological effect known as the
``protégé effect''~\cite{Chase2009}: the teacher feels responsible for his student, commits
to the student's success and possibly experiences student's failure as his own
failure to teach. Teachable computer-based agents have previously been used to
encourage this ``protégé effect'', wherein students invest more effort into
learning when it is for a teachable agent than for themselves~\cite{Chase2009}.
We rely on this cognitive mechanism to reinforce the child's commitment into the
robot-mediated handwriting activity, and we indeed show sustained child-robot
engagement over extended periods of time (several hours spread of a month).

For these two reasons, our approach is to be distinguished from previous works in
educational robotic. Most of these do not consider the agency induced by the
robot beyond its motivational aspect: playing with an interactive, partially autonomous
device naturally induces some form of anthropomorphizing, which leads to some
level of projected agency, and participate the overall excitement (at least, on
the short-term, before the novelty effect vanishes).

In our case, the role of agency is stronger: it induces meta-cognition (``I am
interacting with an agent, so I need to reflect on how to best teach him'')
which is beneficial for the learning process; it also induces a protégé
effect (``I want my robot-agent to succeed!'') which support the commitment of
child into the interaction, also for longer period of time.

@@

We then present and report on the field experiments that we conducted over the
last two years, including four studies at school, one longer experiment
with eight children at the surgery of an occupational therapist, and two
one-month long case studies. While the focus of the school
experiments was mostly technical validation and data acquisition, the three
other experiments involved children with actual deficits, and gave us initial insights on the
relevance and effectiveness of our approach.

@@

A tactile tablet (with a custom application) is used for both the robot and the
child to write: during a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and push the tablet
towards the robot. The robot writes on the tablet by gesturing the writing -- the
letters being actually drawn by the tablet application itself. The child then pull back the
tablet, corrects the robot's attempt by writing him/herself on top or next to
the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her
demonstration to the robot by pressing a small button on the tablet. The robot
``learns'' from this demonstration and tries again.

Since the children are assumed to take on the role of the teachers, we had to ensure
they would be able to manage by themselves the turn-taking and the overall
progression of the activity (moving forwards to the next letter or word). In our design,
the turn-taking relies on the robot prompting for feedback once it is done with
its writing (through simple sentences like ``What do you think?''), and pressing on a
small robot icon on the tablet once the child has finished correcting. In our
experiments, both were easy to grasp for children.

@@

By changing the eigenvalues associated to these eigenvectors and regenerating
letters with the reverse procedure, it becomes then easy to generate new
letters, with distortions that are actually plausible handwriting errors: they
are actually exaggerations of variance of writing styles that naturally
occurs amongst adult writers.  Figure~\ref{fig:sampleLetters} shows examples of
deformed ``g'' generated with such a technique.

@@

\begin{itemize}
    \item the different allographs can by clustered (with a $k$-means or
        mean-shift algorithm) by their visual styles,
    \item we can compute an euclidian distance to the reference letter to assess
        the visual proximity of the demonstration with the expected letter, thus
        providing a quantitative metric of writing performance.
\end{itemize}

@@

The algorithm for machine-learning becomes then a simple matter of converging at
a specific pace towards the child demonstration in the eigenspace.
Figure~\ref{learning_6_demos} (p.~\pageref{learning_6_demos}) illustrates the
process with the complete learning cycle of the number ``6''.

@@

The overall architecture of the system (Figure~\ref{fig:archi}) is therefore
spread over several devices: the {\sc nao} robot itself, that we address via
both a ROS API\footnote{The ROS documentation for {\sc nao} is available at
\url{http://wiki.ros.org/nao}.} and the Aldebaran-provided NaoQI API, one
to four Android tablets (the main tablet is used to draw the robot's letter and
to acquire the children's demonstrations; more tablets have been used in some
studies, either to let the child input words to be written, or for the
experimenter to qualitatively annotate the interaction in a synchronized
fashion), and a central laptop running the machine learning algorithms, the
robot's handwriting gesture generation (based on the NaoQI inverse kinematics
library) and high level control of the activity.

@@

Most of the nodes are written in Python, and the whole source code of the
project is available online\footnote{The primary repository is
\url{https://github.com/chili-epfl/cowriter_letter_learning}.}. Further details
on the technical implementation are available in~\cite{hood2015when}.

@@

The system has been deployed and tested in several situations: in three
different schools (relatively short child-robot interactions with
more than 70 children, aged 5 to 8), in the surgery of an occupational
therapist (with 8 children, each interacting several hours with the system),
and in two case studies that each lasted several weeks. Table~\ref{studies}
gives an overview of these studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}        & {\bf Type}                               & {\bf Avg. duration}    & {\bf Children \#}                     & {\bf Ages} \\ \midrule
{\it School A (1)} & Unstructured group interaction at school & 16 min/group           & 4 $\times$ 8 children                 & 6-7        \\
{\it School B}     & Individual/Pair interaction at school    & 11 min/group           & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it School C}     & Pair interaction at school               & 26 min/group           & 7 $\times$ 2                          & 5-6        \\
{\it School A (2)} & Individual interaction at school         & 20 min                 & 6                                     & 5-6        \\ \midrule
{\it Surgery}      & Individual interaction at surgery        & 3 sessions $\times$ 1h & 8                                     & 6-8        \\ \midrule 
{\it Vincent}      & Case-study, spanning over 4 weeks        & 4 weeks $\times$ 1.5h  & 1                                     & 6          \\
{\it Thomas}       & Case-study, spanning over 4 weeks        & 4 weeks $\times$ 1h    & 1                                     & 5          \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
    \centering
    \includegraphics[height=3.9cm]{schools}
    \includegraphics[height=3.9cm]{schools2}
    \caption{\small The first field studies were focused on technical validation, with
    more than 70 pupils interacting with the robot over short periods (around 15
minutes), either alone or in small groups.}
    \label{fig:schools}
\end{figure}

Critically, these studies were conducted with \emph{whole classes}: we decided
not to select specifically underperforming children as having more children (73
in total) was beneficial for these preliminary studies, and this would have
besides required careful organization with the school due to ethical concerns.

@@

From a technical perspective, the system achieves an acceptable level of
reliability and allow a technically sound autonomous interaction. For instance,
during the second school study (\textit{School B}), the robot withstood
interactions which lasted for a total of 160 minutes.  During this time the
robot wrote 335 letters, 152 of which in response to demonstrations received
from the 21 children. Technical intervention was only required for the three
instances that the robot fell during that day.  Otherwise, the technical
components of the system operated autonomously and as expected over the
sessions.

Due to the modular software architecture (about 10 ROS nodes), the occasional
crashes occurring during others sessions were usually quickly resolved by
re-launching the faulty node alone, and did not significantly impact the
interaction.

The otherwise technical limitations were related to some letters or writing
style (most notably, the ones requiring multiple strokes per letter) not being
adequately processed by the learning algorithm. Support for such letters has
been added as a follow-up of the studies.

@@

More generally, the approach was well accepted and recognized as useful and
promising by teachers and parents; this was \textit{a posteriori} confirmed by
the numerous spontaneous contacts made by parents and therapists who were
looking forward to use the system with their children.

@@

As indicated in the introduction, the literature suggests that 20 handwriting
practice sessions is found to be the minimum to demonstrate effective results in
handwriting remediation~\cite{Hoy2011}. This highlights the necessity to sustain
a child-robot engagement over the long-term if we want to achieve measurable learning gains.

@@

    \begin{tabular}{p{1cm}cccccccc}
        \toprule
        Child & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
        \midrule
        $\mathcal{W}$ & 79.4\% & 81.6\%  & 90.5\% & 87.9\% & 90.7\% & 80.9\% & {\bf 85.2\%} & {\it 5.1} \\ 
        \bottomrule
    \end{tabular}
    \label{tab:results-with-me-ness}
\end{table}

We also conducted a quantitative assessment of the engagement levels of the
children. Table~\ref{tab:results-with-me-ness} reports the levels of
\emph{with-me-ness} of the children during the second study at school A.
\emph{With-me-ness} is a quantifiable precursor of engagement: it measures the
percentage of time spent by the child focusing onto the task at end. It was
first devised in the context of computer-supported
learning~\cite{sharma2014me}, and we have previously studied its applicability to
human-robot interaction and formalized the exact methodology
in~\cite{lemaignan2016realtime}\footnote{Note that with-me-ness is a metric that
allows real-time computation by the robot itself: while we did not yet make
use of it, it would allow the robot to detect possible disengagements, and
eventually address them by adapting its physical behaviour, switching to
different activities, etc.}.

@@

The other three experiments (the group study at the surgery and the two
case-studies) provide further qualitative evidence of engagement over longer
interaction periods. In particular, as reported hereafter, the two 4-weeks long
case studies that have been conducted so far show that our system can sustain
children' engagement over durations (5 hours) that are closer to what is
expected to have an impact in a real therapeutic context.

@@

\begin{inparaenum}[\itshape 1\upshape)]
This experiment was conducted to study \item how easily children with actual
deficit take on the role of a teacher, and \item if they engage \emph{seriously}
in this role. As such, measuring actual handwriting improvements was not a primary goal
of this experiment.
\end{inparaenum}

The experiment took place at the surgery of an occupational therapist in
Normandy, France. Eight children participated, selected by the occupational
therapist based on their age and type of deficit (related to handwriting).
Valérie (female, 7 years old), Antoine (male, 6.5) and Johan (male, 7) are under
the care of the occupational therapist. Émilien (male, 8) and Mathieu (male, 7)
are repeating their school year because of writing difficulties. Marie (female,
6) and Adèle (female, 8) are both ranked at the bottom of their respective
classes in writing activities. Nicolas (male, 7) is under the care of a
neurologist, and has been diagnosed with specific language impairment. Given
their school level, all those children would be expected to know how to
correctly shape cursive letters.

Over a period of two weeks, each child attended three time a one hour long
session (except for Marie and Adèle who only attended one session). The
experimenter role was limited to the explanation of the rules of the game and
the basic tablet usage. For this experiment, the children were provided with two
tablets: one to choose the words (or letters) to teach to the robot, the other one
to write, like in the previous studies. We also provided templates of the
cursive letters, where the children to ask for it.

We only provided the children with minimal explanations on the task (they would
have to help the robot to improve its writing style), so as to assess if and how
children would naturally take on the role of the teacher, and how seriously they
engage into helping the robot.

This assessment is carried through two additional buttons on the tablet: a
green ``thumbs up'' and a red ``thumbs down''. The children are told to use
them to evaluate the robot's improvements (``thumb up'' being a way to give
positive feedback while ``thumb down'' would convey negative feedback). Our
assessment builds on the hypothesis that the more the children provides feedback to
the robot with the green/red buttons, the more they assume the role of a
teacher. Then, by correlating the feedback with the actual performance of the
robot, we can measure to what extend the children are seriously assuming their
role: since the child knows what a correct letter should look like, if their
feedback does not correlate with the actual performance of the robot, the child
is likely engaged into a \emph{playful} instead of \emph{serious} role-playing.

@@

\begin{table}
    \centering
    \begin{tabular}{ccccll}
        \toprule
        Child          & \#demos & \#positive & \#negative & $r_{robot}$ & \\ \midrule
        \emph{Valérie} & 127 & 24 & 6 & 0.25 \small\tt **  \\ 
        \emph{Émilien} & 223 & 20 & 9 & 0.06 \small\it ns  \\
        \emph{Mathieu} & 131 & 10 & 3 & 0.23 \small\tt **  \\
        \emph{Nicolas} & 115 & 16 & 4 & 0.31 \small\tt *** \\
        \emph{Johan}   & 98 & 10 & 5  & 0.10 \small\it ns  \\
        \emph{Antoine} & 83 & 10 & 3  & 0.20 \small\tt *   \\
        \emph{Adèle}   & 35 & 4 & 2   & 0.28 \small\tt *   \\
        \emph{Marie}   & 40 & 5 & 1   & -0.02 \small\it ns \\ \midrule
        Total          & 852 & 99 & 33  &   & \\ \bottomrule
    \end{tabular}
    \caption{\small Feedback from the children to the robot. \emph{\#demos}
        denotes the total number of demonstrations drawn by the children;
        \emph{\#positive} and \emph{\#negative} the number of positive \resp
        negative feedbacks they provided. $r_{robot}$ reflects the correlation
        between the feedback provided by the children and the performance of the
        robot. Results taken from~\cite{jacq2016building}.}

@@

All children maintained their engagement during the entire sessions. They
provided on average 42 demonstrations per session. All children did use the
feedback buttons (in total, 99 ``thumb up'' and 33 ``thumb down'' were recorded,
see Table~\ref{table:scores}).

To study the correlation between the children's feedback and the robot's
actual improvements, we estimate the robot's progress as the difference between an
initial score (average euclidian distance between the shape drawn by the robot
at its first attempt and the shape of the reference letter) and the robot's
score at the current round of demonstrations. We then correlate this progress
to the positive or negative feedback provided by the children, following a
statistical method presented in~\cite{jacq2016building}.

We find (Table~\ref{table:scores}) that the feedback of five out of the eight
children do significantly correlate with the actual performance of the robot:
this indicates that these children are effectively taking on the teacher's
clothes and are seriously providing feedback to the robot. \fixme{what can we
say about the 3 others? they did not play seriously, or they were unable to
actually recognize the robot mistakes?}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{diego}
    \caption{\small Vincent correcting {\sc nao}'s attempt by rewriting the
        whole word. Empty boxes are drawn on the screen to serve as template for the child
        and to make word segmentation more robust.}
    \label{fig:diego}
\end{figure}

@@

To complement the intrinsic motivation of helping a robot to communicate with another one, we
gradually increased the complexity of Vincent's task to keep it challenging and
interesting (the first week: demonstration of single letters; the second week:
short words; the third week: a full letter -- Figure~\ref{fig:stimuli}).

@@

    \caption{\small (French) text generated by the robot, before and after a one
        hour long interaction session with the child. The red box
        highlights one instance of striking improvement of the robot's
        handwriting legibility.}

@@

Over the whole duration of the study, Vincent had provided 154 demonstrations to the robot, and he remained
actively engaged over the four weeks. The story was well accepted by the child and
he seriously engaged into the game. After the first week, he showed good
confidence to play with the robot and by the end of the study he had built
affective bonds with the robot, as evidenced by several letters he did send to the
robot \emph{after} the end of the study (one of them four months later) to get
news. This represents an initial validation of our hypothesis: our system can effectively keep a
child engaged with the robot for a relatively long periods of time (about 5
hours spread over a month).

No hard conclusion can be drawn in terms of actual handwriting remediation: we did
not design this study to formally assess possible improvements. However, as
visible on Figure~\ref{fig:stimuli}, Vincent was able to significantly improve
the robot's skill, and post-hoc interviews revealed that he acknowledged that he
was the one helping the robot. In that respect, Vincent realized that he was
``good enough'' at writing to help someone else. Quoting the mother a week
after the end of the experiment: ``Vincent's handwriting has changed over the
last weeks, going from a mix of standalone and cursive letters to full words in
cursive. This requires a lot of efforts and concentration from him, but he did
succeed during the sessions with the robot as he knew he had to show a
consistent writing to the robot''.

>>>>

\author{Séverin Lemaignan$^1$, Alexis Jacq$^{1,3}$, Deanna Hood$^1$, Fernando Garcia$^1$, \\
    Aude Billard$^2$, Ana Paiva$^3$, Pierre Dillenbourg$^1$ \\
$^1$CHILI Lab, École Polytechnique Fédérale de Lausanne, Suisse,\\
$^2$LASA, École Polytechnique Fédérale de Lausanne, Suisse,\\
$^3$Instituto Superior Técnico, University of Lisbon, Portugal}

@@

Thomas is five and half, and has been diagnosed with visuo-constructive
deficits. He is under the care of an occupational therapist, and tries to
workaround his inability to draw letters in a consistent manner. Vincent is six
and struggles at school with his poor handwriting and even poorer
self-confidence.\footnote{All children's names have been changed.}

@@

Remediations for handwriting difficulties traditionally involve long
interventions (at least 10 weeks~\cite{Hoy2011}), essentially consisting in
handwriting training with occupational therapists, and primarily addressing the
\emph{motor} deficits.  Improvements in self-confidence and anxiety occur (at
best) as a side-effect of the child improving their handwriting skills and,
consequently, improving their performance at school.

We present in this article a new take on this educative challenge: a remediation
procedure that involve a ``bad writer'' robot that is \emph{taught} by the
child. By building on the \emph{learning by teaching} paradigm, not only the
child practises handwriting but, as they take on the role of the teacher,
they also positively reinforce their self-esteem and motivation: their
social role shift from the ``underperformer'' to ``the one who knows and
teaches''. And by relying on a robot, we can tailor the exercises and the
learning curve to each children' needs, as we will show in this article.

\subsection{Learning by Teaching Handwriting}

@@

We also investigate here a particular role for a robot in the
education of handwriting: not only is the robot actively performing the activity
by drawing letters, but it does so in a way that engages the child in a very
specific social role. The child is the teacher in this relationship and the
robot is the learner: the child is to engage in a (meta-) cognitive relationship
with the robot to try to understand why the robot fails and how to help it best.
Here, the robot is more than just an activity facilitator or orchestrator -- its
physical presence and embodiment induce agency and anthropomorphising, and
cognitively engage the child into the learning activity (be it consciously or
not).

The commitment of the child into the interaction build on a
psychological effect known as the
``protégé effect''~\cite{Chase2009}: the teacher feels responsible for his student, commits
to the student's success and possibly experiences student's failure as his own
failure to teach. Teachable computer-based agents have previously been used to
encourage this ``protégé effect'', wherein students invest more effort into
learning when it is for a teachable agent than for themselves~\cite{Chase2009}.
We rely on this cognitive mechanism to reinforce the child's commitment into the
robot-mediated handwriting activity, and we indeed show sustained child-robot
engagement over extended periods of time (several hours spread of a month).

For these two reasons, our approach is to be distinguished from previous works in
educational robotic. Most of these do not consider the agency induced by the
robot beyond its motivational aspect (playing with an interactive, partially autonomous
device naturally induces some form of anthropomorphizing, which leads to some
level of projected agency, and participate the overall excitement -- at least, on
the short-term, before the novelty effect vanishes).
In our case, the role of agency is stronger: it induces meta-cognition (``I am
interacting with an agent, so I need to reflect on how to best teach him'')
which is beneficial for the learning process; it also induces a protégé
effect (``I want my robot-agent to succeed!'') which support the commitment of
child into the interaction, also for longer period of time.

@@

We then present and report on the field experiments that we have conducted over the
last two years, including four studies at school, one longer experiment
with eight children at the surgery of an occupational therapist, and two
one-month long case studies. While the focus of the school
experiments was mostly the technical validation and data acquisition, the three
other experiments involved children with actual deficits, and gave us initial insights on the
relevance and effectiveness of our approach.

@@

A tactile tablet (with a custom application) is used for both the robot and the
child to write. During a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and push the tablet
towards the robot. The robot writes on the tablet by gesturing the writing in
the air,  the letters being actually drawn by the tablet application itself. The child then pull back the
tablet, corrects the robot's attempt by writing him/herself on top or next to
the robot's writing (see Figure~\ref{fig:diego}), and ``send'' his/her
demonstration to the robot by pressing a small button on the tablet. The robot
``learns'' from this demonstration and tries again.

Since the children are assumed to take on the role of the teachers, we had to ensure
they would be able to manage by themselves the turn-taking and the overall
progression of the activity (moving forwards to the next letter or word). In our design,
the turn-taking relies on the robot prompting for feedback once it is done with
its writing (through simple sentences like ``What do you think?''), and pressing on a
small robot icon on the tablet once the child has finished correcting. In our
experiments, and once introduced by the experimenter, both were easy to grasp
for all the children.

@@

By changing the eigenvalues associated to these eigenvectors and regenerating
letters with the reverse procedure, it becomes then easy to generate new
letters, with distortions that are actually plausible handwriting errors: they
are exaggerations of variance of writing styles that naturally
occurs amongst adult writers.  Figure~\ref{fig:sampleLetters} shows examples of
deformed ``g'' generated with such a technique.

@@

\begin{itemize}
    \item the different allographs can by clustered (with a $k$-means or
        mean-shift algorithm) by their visual styles,
    \item we can compute an euclidian distance to the reference letter to assess
        the topological proximity of the demonstration with the expected letter, thus
        providing a quantitative metric of writing performance.
\end{itemize}

@@

The algorithm for machine-learning becomes then a simple matter of converging at
a specific pace towards the child demonstration in the eigenspace.
Figure~\ref{learning_6_demos} (p.~\pageref{learning_6_demos}) illustrates the
process with a complete learning cycle of the number ``6''.

@@

The overall architecture of the system (Figure~\ref{fig:archi}) is therefore
spread over several devices: the {\sc nao} robot itself, that we address via
both a ROS API\footnote{The ROS documentation for {\sc nao} is available at
\url{http://wiki.ros.org/nao}.} and the Aldebaran-provided NaoQI API, one
to four Android tablets (the main tablet is used to draw the robot's letter and
to acquire the children's demonstrations; more tablets have been used in some
studies, either to let the child input words to be written, or for the
experimenter to qualitatively annotate the interaction in a synchronized
fashion), and a central laptop running the machine learning algorithms, the
robot's handwriting gesture generation (based on the NaoQI inverse kinematics
library) and the high-level control of the activity (relying on {\sc
pyRobots}~\cite{lemaignan2015pyrobots} and a custom finite state machine).

@@

Most of the nodes are written in Python, and the whole source code of the
project is available online\footnote{The primary repository is
\url{https://github.com/chili-epfl/cowriter_letter_learning}.}. The details
of the technical implementation are available in~\cite{hood2015when}.

@@

The system has been deployed and tested in several situations: in three
different schools (more than 70 children aged 5 to 8; relatively short duration interactions), in the surgery of an occupational
therapist (8 children; each interacting several hours with the system),
and during two case studies that each lasted several weeks. Table~\ref{studies}
gives an overview of these studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}        & {\bf Type}                               & {\bf Avg. duration}    & {\bf \# Children}                     & {\bf Ages} \\ \midrule
{\it School A (1)} & Unstructured group interaction at school & 16 min/group           & 4 $\times$ 8 children                 & 6-7        \\
{\it School B}     & Individual/Pair interaction at school    & 11 min/group           & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it School C}     & Pair interaction at school               & 26 min/group           & 7 $\times$ 2                          & 5-6        \\
{\it School A (2)} & Individual interaction at school         & 20 min                 & 6                                     & 5-6        \\ \midrule
{\it Surgery}      & Individual interaction at surgery        & 3 sessions $\times$ 1h & 8                                     & 6-8        \\ \midrule 
{\it Vincent}      & Case-study                               & 4 weeks $\times$ 1.5h  & 1                                     & 6          \\
{\it Thomas}       & Case-study                               & 4 weeks $\times$ 1h    & 1                                     & 5          \\ \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
    \centering
    \includegraphics[height=3.9cm]{schools}
    \includegraphics[height=3.9cm]{schools2}
    \caption{\small The first field studies were focused on technical validation, with
    more than 70 pupils interacting with the robot over short periods (between
    10 and 25 minutes), either alone or in small groups.}
    \label{fig:schools}
\end{figure}

Critically, these studies were conducted with \emph{whole classes}: we decided
not to select specifically underperforming children as having more children (73
in total) was beneficial for these preliminary studies. Besides, due to ethical concerns, this would have
required complex organization with the school that we wanted to avoid at the validation stage.

@@

From a technical perspective, the system achieves an acceptable level of
reliability and allow a technically sound autonomous interaction. For instance,
during the second school study (\textit{School B}), the robot withstood
interactions which lasted for a total of 160 minutes.  During this time the
robot wrote 335 letters, 152 of which in response to demonstrations received
from the 21 children. Technical intervention was only required for the three
instances that the robot fell during that day.  Otherwise, the technical
components of the system operated autonomously and as expected with all the groups of children.

Due to the modular software architecture (mostly independent ROS nodes), the occasional
crashes occurring during others studies were usually quickly resolved by
re-launching the faulty node alone, and did not significantly impact the
interaction.

The otherwise technical limitations were related to some letters or writing
style (most notably, the ones requiring multiple strokes per letter) not being
adequately processed by the learning algorithm. Support for such letters was
added as a follow-up of the validation studies.

@@

During two focus groups organized with teachers, the welcomed the approach and
recognized it as useful and
promising; this was \textit{a posteriori} confirmed by
multiple spontaneous contacts made by parents and therapists who were
looking forward to use the system with their children.

@@

The literature suggests that 20 handwriting
practice sessions is found to be the minimum to demonstrate effective results in
handwriting remediation~\cite{Hoy2011}. This highlights the necessity to sustain
a child-robot engagement over the long-term if we want to achieve measurable learning gains.

@@

    \begin{tabular}{p{1cm}cccccccc}
        \toprule
        \bf Child & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
        \midrule
        $\mathcal{W}$ & 79.4\% & 81.6\%  & 90.5\% & 87.9\% & 90.7\% & 80.9\% & {\bf 85.2\%} & {\it 5.1} \\ 
        \bottomrule
    \end{tabular}
    \label{tab:results-with-me-ness}
\end{table}

We also conducted a quantitative assessment of the engagement levels of the
children. Table~\ref{tab:results-with-me-ness} reports the levels of
\emph{with-me-ness} of the children during the second study at school A.
\emph{With-me-ness} is a quantifiable precursor of engagement: it measures the
percentage of time spent by the child focusing onto the task at end. This metric was
first devised in the context of computer-supported
learning~\cite{sharma2014me}, and we have previously studied its applicability to
human-robot interaction and formalized the exact methodology
in~\cite{lemaignan2016realtime}\footnote{Note that with-me-ness is a metric that
allows real-time computation by the robot itself: while we did not yet make
use of it, it does in principle allow the robot to detect on-line possible disengagements, and
eventually address them by adapting its physical behaviour, switching to
different activities, etc.}.

@@

The other three experiments (the group study at the surgery and the two
case-studies) provide further qualitative evidence of engagement over longer
interaction periods. In particular, as reported hereafter, the two 4-weeks long
case studies that have been conducted so far show that our system can sustain
children' engagement over durations (5 hours) that are getting closer to what is
expected to have an impact in a real therapeutic context.

@@

\begin{inparaenum}[\itshape 1\upshape)]
This experiment was conducted to study \item how easily children with actual
deficits take on the role of a teacher, and \item if they engage \emph{seriously}
in this role. Measuring actual handwriting improvements was not a primary goal
of this experiment.
\end{inparaenum}

The experiment took place at the surgery of an occupational therapist in
Normandy, France. Eight children participated, selected by the occupational
therapist based on their age and type of deficit (all related to handwriting).
Valérie (female, 7 years old), Antoine (male, 6.5) and Johan (male, 7) are under
the direct care of the occupational therapist. Émilien (male, 8) and Mathieu (male, 7)
are repeating their school year because of writing difficulties. Marie (female,
6) and Adèle (female, 8) are both ranked at the bottom of their respective
classes in writing activities. Nicolas (male, 7) is under the care of a
neurologist, and has been diagnosed with specific language impairment. Given
their age and school year, all those children would be expected to know how to
correctly shape cursive letters.

Over a period of two weeks, each child attended three time a one hour long
session (except for Marie and Adèle who only attended one session). The
experimenter role was limited to the explanation of the task and
the basic tablet usage. For this experiment, the children were provided with two
tablets: one to choose the words (or letters) to teach to the robot, the other one
to write, like in the validation studies. We also provided paper-based templates of the
cursive letters, where the children to ask for them.

We only provided the children with minimal explanations on the task (they would
have to help the robot to improve its writing style), so as to assess if and how
children would naturally take on the role of the teacher. We besides assessed how seriously they
engage into helping the robot through two additional buttons on the tablet: a
green ``thumbs up'' and a red ``thumbs down''. The children were told to freely use
them to evaluate the robot's improvements (``thumb up'' to give
positive feedback, ``thumb down'' to convey negative feedback). Our
assessment builds on the hypothesis that the more the child provides feedback to
the robot, the more they assume the role of a
teacher. Then, by correlating the feedback with the actual performance of the
robot, we can measure to what extend the children are seriously assuming their
role: if their feedback does correlate with the actual performance of the robot, the child
is likely engaged into a \emph{serious} teaching role-playing.

@@

\begin{table}
    \centering
    \begin{tabular}{ccccl}
        \toprule
        \bf Child      & \bf \# Demos/hour & \bf \# Positive & \bf \# Negative & $r$ \\ \midrule
        \emph{Valérie} & 42           & 24              & 6               & 0.25 \small\tt **  \\ 
        \emph{Émilien} & 74           & 20              & 9               & 0.06 \small\it ns  \\
        \emph{Mathieu} & 43           & 10              & 3               & 0.23 \small\tt **  \\
        \emph{Nicolas} & 38           & 16              & 4               & 0.31 \small\tt *** \\
        \emph{Johan}   & 32           & 10              & 5               & 0.10 \small\it ns  \\
        \emph{Antoine} & 27           & 10              & 3               & 0.20 \small\tt *   \\
        \emph{Adèle}   & 35           & 4               & 2               & 0.28 \small\tt *   \\
        \emph{Marie}   & 40           & 5               & 1               & -0.02 \small\it ns \\ \bottomrule
    \end{tabular}
    \caption{\small Feedback from the children to the robot. \emph{\#Demos}
        denotes the average number of demonstrations per hour provided by the children;
        \emph{\#Positive} and \emph{\#Negative} the total number of positive \resp
        negative feedbacks they provided. $r$ is the correlation coefficient
        between the feedback provided by the children and the performance of the
        robot. Results taken from~\cite{jacq2016building}.}

@@

All children maintained their engagement during the entire sessions. They
provided on average 42 demonstrations per session. All children did use the
feedback buttons (in total, 99 ``thumb up'' and 33 ``thumb down'' were recorded,
see Table~\ref{table:scores}). This indicates that they are all able to engage into playing
the role of a teacher.

To study the correlation between the children' feedback and the
actual improvements of the robot, we estimate the robot's progress as the difference between an
initial score (average euclidian distance between the shape drawn by the robot
at its first attempt and the shape of the reference letter) and the robot's
score at the current round of demonstrations. We then correlate this progress
to the positive or negative feedback provided by the children (details of the
method are presented in~\cite{jacq2016building}).

We find that the feedback of five out of the eight
children do significantly correlate with the actual performance of the robot (Table~\ref{table:scores}).
This indicates that these children are effectively taking on the teacher's
clothes and are seriously providing feedback to the robot. The
observation of the three remaining children reveal a variety of behaviours (for
instance, one was actually rating how ``nice'' the robot is, and another one
had rather ``artistic'' writing style preferences that were independent of the
actual legibility of the robot), and during this specific experiment, we eventually
found that they were all seriously engaged into teaching.

To summarize, this experiment shows that children with actual handwriting impairments
do accept well the robot, can commit themselves into long interactions, do take
on the role of the teacher (as evidenced by both the number of writing demonstrations
they provided and their self-inclination to give feedback to the robot), and
finally take their teacher's role seriously.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{diego}
    \caption{\small Vincent correcting {\sc nao}'s attempt by rewriting the
        whole word. Empty boxes are drawn on the screen to serve as template for the child
        and to make letter segmentation more robust.}
    \label{fig:diego}
\end{figure}

@@

To supplement the intrinsic motivation of helping a robot to communicate with another one, we
gradually increased the complexity of Vincent's task to keep it challenging and
interesting (the first week: demonstration of single letters; the second week:
short words; the third week: a full letter -- Figure~\ref{fig:stimuli}).

@@

    \caption{\small Text (in French) generated by the robot, before and after a one
        hour long interaction session with the child. The red box
        highlights one instance of striking improvement of the robot's
        handwriting legibility.}

@@

Over the whole duration of the study, Vincent had provided 154 demonstrations to the robot, and he remained
actively engaged over the four weeks. The story was well accepted by the child and
he seriously engaged into the game. After the first week, he showed good
confidence to play with the robot and by the end of the study he had built
affective bonds with the robot, as evidenced by several letters he did send to the
robot \emph{after} the end of the study (one of them four months later) to get
news. This represents an initial validation of our hypothesis: our system can effectively keep a
child engaged with the robot for a relatively long period of time (about 5
hours spread over a month), and we can build a tutor/protégé relationship.

No hard conclusion can be drawn in terms of actual handwriting remediation as we did
not design this study to formally assess possible improvements. However, as
visible on Figure~\ref{fig:stimuli}, Vincent was able to significantly improve
the robot's skill, and  he acknowledged that he
was the one helping the robot during post-hoc interviews. In that respect, Vincent realized that he was
``good enough'' at writing to help someone else. The fact that Vincent seriously
committed to the role of a teacher is further supported by feedback sent by
Vincent's parents a week after the end of the experiment: ``Vincent's handwriting has changed over the
last weeks, going from a mix of standalone and cursive letters to full words in
cursive. This requires a lot of efforts and concentration from him, but he did
succeed during the sessions with the robot as he knew he had to show a
consistent writing to the robot''.
